Møil: A web-based administration tool for mail servers backed by relational databases

### 2014-07-30 doc/study/study.tex
Finished comparison overview table and summaries for contestants.

\usepackage{url}
%\urldef{\urlTcpdump}{\url}{http://www.tcpdump.org/}
\urldef{\urlARSupport}{\url}{http://api.rubyonrails.org/classes/ActiveRecord/Migration.html#class-ActiveRecord::Migration-label-Database+support}
\urldef{\urlPHPBashing}{\url}{http://eev.ee/blog/2012/04/09/php-a-fractal-of-bad-design}

@@

			\subsubsection*{Data model (1)}
				From a database engineering perspective, it is in general
				favourable to design a database with a \crit{normalized
				relational database schema}{1.1} \cite{dbnorm}. For a query to
				check the existence of a mailbox for example, a JOIN or nested
				query would be necessary, which has a very small performance
				impact. However, the schema normalization prevents data
				redundancies and inconsistencies. \emph{postfix.admin} does not
				have a normalized schema, so any solution which has one can not
				be \crit{compatible to \emph{postfix.admin}}{1.2}. A
				compatibility could be useful, if there is a big database
				already administered with \emph{postfix.admin}, which shall be
				migrated to another tool. Another possibility for this is an
				\crit{import of \emph{postfix.admin} data}{1.3}. When it comes
				to updates to the administration interface, \crit{means to
				handle database schema updates}{1.4} ease the update process
				(and therefore increase the probability of the actual
				conduction of an update). Last but not least, \crit{database
				independence}{1.5} allows a free choice of the \ac{DBMS} and
				integration with other software.

@@

			\subsubsection*{Security \& robustness (3)}
				For now, only three criteria in this category shall be
				considered, because further security analysis would require an
				in-depth source code review. The \crit{password hashing}{3.1}
				availability and algorithm strength is important against
				unauthorized database access through potential security holes
				in the application. When the database is maliciously read, the
				value of the passwords to the attacker is inversely
				proportional to the time it costs, to crack the password
				hashing. \crit{Test coverage}{3.2} is important for this
				evaluation and for security purposes, because it lowers the
				probability of bugs per lines of code by trend. As
				\crit{Containability}{3.3}, the modular independence (e.g.
				independence of the web interface from the Message Store file
				system) from a security perspective is perceived.

			\subsubsection*{Responsive design (UI) (4)}
				In the days of rapidly increased mobile device usage, a
				responsive design (e.g. different layout for different screen
				sizes) is an important attribute of user interfaces. Usually,
				responsive design means \crit{adapting to mobile devices}{4.1}
				as phones and tablet computers. In this document, also the
				\crit{ability of usage from a \ac{CLI}}{4.2} is considered,
				because it is widespread among system administrators.

@@

				\emph{modoboa} seems the most mature of the compared tools. It
				uses the Python Django framework and therefore includes
				facilities like schema migrations.

				% Criteria
				% 	2.4.1	No undo.
				%   4.1		Bad mobile experience.

@@

%				As said, \emph{postfix.admin} seems to be a (negative) example
%				not only for \emph{Møil}. Many solutions aim to overcome its
%				shortcomings. \emph{postfix.admin} uses a non-normalized
%				\cite{dbnorm} database schema, which could spare a JOIN in some
%				database queries but is very prone to inconsistencies. There
%				are no means of handling database schema updates; the schema is
%				imported via a \ac{SQL} file on the initial installation. Also
%				it creates an alias entry for every mailbox (even when the mail
%				is not forwarded). This may have a cause, but is messy anyway.

				% Criteria
				%	1.4		Migrations one big file upgrade.php with functions
				%			like upgrade_<n>_<db>.
				%	1.5		No systematic database independence (or object
				%			relational mapper at all), only manual support for
				%			mysql and pgsql.
				%	2.4.1	No undo.
				%	3.1		Password hash default md5-crypt.
				%	3.3		Can move mail data if configured and access on Mail
				%			Store, but disabled by default.

			\subsubsection*{PostVis Admin}
				\infotable{http://postvisadmin.sourceforge.net}{PHP}{2010-06-24}{183}
				% SQL injections
				% Badly strutured
				% Code duplication
				% Best example for non-PHP policy
				%
				% Criteria
				%	1.2		Also not normalized like postfix.admin and much
				%			table names and fields equally named.
				%	2.3		Even on a quick view, access control does not seem
				%			effective. Any logged in user (also non-superadmin
				%			users) seem to be able to delete any other user.
				%	3.1		Default md5-crypt.
				%	3.3		Features do not work without shell execution on same
				%			server (e.g. amavisd-release).

@@

				The Development website was not linked correctly; the linked
				hostname did not even exist in DNS, so the URL of the VCS
				could not be determined and the 0.3-r2 version was examined.

				% Bad struture (almost everything inside include/libratuus.php)
				% but way more organized (object oriented, systematic 'routing'
				% and access control) than PostVis Admin, for example. No query
				% sanitization in mailClass->ask_db(). mysql_query() supports
				% multiple queries configuration dependently.
				%
				% Criteria
				%	2.4.1	No undo.
				%	3.1		Only md5-crypt.

@@

				% Manual SQL queries, but sanitized. Not easy to get
				% information as for example, if it is database independent.
				%
				% Criteria
				% 	1.5		MySQL dependend in some places (w.g. in maintainance
				%			code), independend in others (e.g. authentication).
				%	3.2		Just custom SMTP proxy for spam maintainance
				%			integration seems to be tested a little.

@@

				% PHP und ORM!!
				% Criteria
				%	2.4.1	No undo.
				%	2.4.3	Support for access restriction for mailboxes to
				%			SIEVE.
				%	3.1		Default is salted MD5, "more" secure algorithms
				%			only supported with dovecotpw call.
				%	3.3		See 3.1.

			\begin{landscape}
				\subsubsection*{Comparison overview}
				\label{sec:evaluation:overview}
					% 1. modoboa
					% 2. Møil
					% 3. VBox.Adm?
					% 4. postfix.admin
					% 5. ratuus
					% 6. PostVis Admin

					\begin{tabular}{l|cccccc|ccccccccc}
						              & 1 & 1.1 & 1.2  & 1.3 & 1.4   & 1.5  & 2 & 2.1 & 2.2 & 2.3  & 2.4 & 2.4.1 & 2.4.2 & 2.4.3 & 2.4.4 \\
						\hline
						modoboa       &   & \y  &  \n  & \y  &  \y   &  \y  &   & \y  & \y  &  \y  &     & (\y)  & \y    &  \y   & \y    \\
						Møil          &   & \y  &  \n  & \y  &  \y   &  \y  &   & \y  & \y  &  \y  &     &  \y   & \n    &  \n   & \n    \\
						postfix.admin &   & \n  &      &     & (\y)  & (\y) &   & \y  & \y  &  \y  &     & (\y)  & \y    &  \n   & \n    \\
						PostVis Admin &   & \n  & (\y) & \n  &  \n   &  \n  &   & \y  & \n  & (\y) &     &  \n   & \n    &  \n   & \y    \\
						ratuus        &   & \n  &  \y  & \y  &  \n   &  \n  &   & \y  & \y  &  \y  &     & (\y)  & \n    &  \n   & \n    \\
						VBox.Adm      &   & \y  &  \n  & \y  &  \n   & (\y) &   & \y  & \y  &  \y  &     &  \n   & \y    &  \n   & \y    \\
						ViMbAdmin     &   & \y  &  \n  & \n  &       &  \y  &   & \y  & \y  &  \y  &     & (\y)  & \n    & (\n)  & \n    \\
						\hline
					\end{tabular}
					\s

					\noindent
					\begin{tabular}{l|cccc|ccc|ccc}
						              & 3 & 3.1  & 3.2  & 3.3  & 4 & 4.1  & 4.2 \\
						\hline
						modoboa       &   &  \y  &  \y  &  \y  &   & (\y) & \y  \\
						Møil          &   &  \y  &  \y  &  \y  &   &  \y  & \y  \\
						postfix.admin &   & (\y) &  \n  & (\y) &   &  \n  & \y  \\
						PostVis Admin &   & (\y) &  \n  &  \n  &   &  \n  & \n  \\
						ratuus        &   & (\y) &  \n  &  \y  &   &  \n  & \n  \\
						VBox.Adm      &   &  \y  & (\y) &  \y  &   &  \n  & \y  \\
						ViMbAdmin     &   & (\y) &      & (\n) &   &      &     \\
						\hline
					\end{tabular}
					\s

@@

		That fact alone would maybe suffice as an argument for the development
		for yet another tool but was not as important as some other ideas.
		\emph{Møil} began as a replacement for the database schema shipped with
		\emph{postfix.admin} in a private mail system of mine. The schema
		was intended to be reimplemented normalized \cite{dbnorm} and with
		a possibility of being updated through schema
		migrations\footnote{\url{http://en.wikipedia.org/wiki/Schema_migration}}).
		Also, the \ac{Rails} project was intended to provide the basis for
		command line utilities to import the existent data and to perform basic
		administration tasks on domains, mailboxes and aliases. As there has
		been only the schema in use, not the whole \emph{postfix.admin} web
		frontend, there was an extra solution to generate password hashes and
		send them to an administrative contact in place. A replacement for this
		extra tool at first was the only function of the web interface part of
		\emph{Møil}. The migration opportunity was also used to change the
		password hashing algorithm from unsalted \texttt{md5} to
		\texttt{sha512-crypt}, which is substantially more secure
		\cite{missing}. (\texttt{bcrypt} was considered but unfortunately
		rejected, because it is not supported by \texttt{glibc} included in
		Debian Linux 7.)

@@

	\newpage
	\section*{Numbered index of criteria}
	\label{sec:appendix:criteria}
		\begin{easylist}
			& Data model
			&& Normalized schema
			&& \emph{postfix.admin} compatible
			&& \emph{postfix.admin} import
			&& Updatable (Migrations)
			&& Database independence

>>>>

\usepackage{url}
\urldef{\urlARSupport}{\url}{http://api.rubyonrails.org/classes/ActiveRecord/Migration.html#class-ActiveRecord::Migration-label-Database+support}
\urldef{\urlPHPBashing}{\url}{http://eev.ee/blog/2012/04/09/php-a-fractal-of-bad-design}

@@

			\todo{Sources for most of the criteria!}

			\subsubsection*{Data model (1)}
				From a database engineering perspective, it is in general
				favourable to design a database with a \crit{normalized
				relational database schema}{1.1} \cite{dbnorm}. For a query to
				check the existence of a mailbox for example, a JOIN or nested
				query would be necessary, which has a very small performance
				impact. However, the schema normalization prevents data
				redundancies and inconsistencies. \emph{postfix.admin} does not
				have a normalized schema, so any solution which has one can not
				be \crit{compatible to \emph{postfix.admin}}{1.2}. A
				compatibility could be useful, if there is a big database
				already administered with \emph{postfix.admin}, which shall be
				migrated to another tool. Another possibility for this is an
				\crit{import of \emph{postfix.admin} data}{1.3}. When it comes
				to updates to the administration interface, \crit{means to
				handle database schema updates}{1.4} ease the update process
				(and therefore increase the probability of the actual
				conduction of an update). Last but not least, \crit{systematic
				database independence}{1.5} abstracts from raw \ac{SQL} queries
				and allows a free choice of the \ac{DBMS} and integration with
				other software.

@@

			\subsubsection*{Security \& robustness (3)}
			\todo{Password hashing algorithms strenghts}
				For now, only three criteria in this category shall be
				considered, because further security analysis would require an
				in-depth source code review. The \crit{password hashing}{3.1}
				availability and algorithm strength is important against
				unauthorized database access through potential security holes
				in the application. When the database is maliciously read, the
				value of the passwords to the attacker is inversely
				proportional to the time it costs, to crack the password
				hashing. \crit{Test coverage}{3.2} is important for this
				evaluation and for security purposes, because it lowers the
				probability of bugs per lines of code by trend. As
				\crit{Containability}{3.3}, the modular independence (e.g.
				independence of the web interface from the Message Store file
				system) from a security perspective is perceived.

			\subsubsection*{Responsive design (\acs{UI}) (4)}
				In the days of rapidly increased mobile device usage, a
				responsive design (e.g. different layout for different screen
				sizes) is an important attribute of user interfaces. Usually,
				responsive design means \crit{adapting to mobile devices}{4.1}
				as phones and tablet computers. In this document, also the
				\crit{ability of usage from a \ac{CLI}}{4.2} is considered,
				because it is widespread among system administrators.

@@

				Of the compared tools, \emph{modoboa} seems to be one of the
				most mature ones. It uses the Python Django framework and
				therefore includes facilities like schema migrations and an
				\ac{ORM}. It passes almost every criterium; just when it comes
				to the \crit{audit trail}{2.4.1}, there is no undo support.
				Also the user interface on a mobile phone, was a little sloppy.

@@

				For many of the solutions, \emph{postfix.admin} seems to be an
				inspiration and also the impulse for a new implementation. Many
				solutions aim to overcome its shortcomings, but many also fail
				to live up to that task (which in this case means every PHP
				tool). \emph{postfix.admin} uses a non-normalized \cite{dbnorm}
				database schema, which could spare a JOIN in some database
				queries but is very prone to inconsistencies. \crit{Schema
				migrations}{1.4} are not present through a database abstraction
				library or an \ac{ORM}, but manually implemented rather
				minimalist. There is one big file (\texttt{upgrade.php})
				containing a pile of functions like
				\texttt{upgrade\_<n>\_<db>}, which contain database dependent
				raw \ac{SQL} queries to perform the transformations on the
				schema. There is no \crit{systematic database
				independence}{1.5} or an \ac{ORM}, just "manual" support for
				MySQL and PostgreSQL. Oddly, for every mailbox there is an
				alias entry created (even when the mail is not forwarded).
				\emph{postfix.admin} has support for an \crit{audit
				trail}{2.4.1} but without undo. The default \crit{password
				hashing}{3.1} algorithm is \texttt{md5-crypt}. It is
				\crit{containable}{3.3} by default but does not support the
				movement of mail data, then, which would require the direct
				access to the Mail Store.

			\subsubsection*{PostVis Admin}
				\infotable{http://postvisadmin.sourceforge.net}{PHP}{2010-06-24}{183}

				PostVis Admin is the best acknowledgement for the non-PHP
				policy it was decided on when searching for a
				\emph{postfix.admin} replacement. The code is very bad
				structured and full of redundancies. There is no \crit{schema
				normalization}{1.1} and the schema seems \crit{compatible to
				\emph{postifx.admin}}{1.2} but is likely to depart from it in
				some details. \ac{SQL} injections are findable just by checking
				for criteria like if there is an \crit{access control}{2.3}
				feature, which is basically existing but not very effective,
				because it is not implemented securely. (Any logged in user
				(also non-superadmin users) seem to be able to delete any other
				user.) The default \crit{password hashing}{3.1} algorithm is
				\texttt{md5-crypt} and there are features, which do not work
				without shell execution privileges on the mail system servers
				(e.g. \texttt{amavisd-release}, so the solution is not really
				\crit{containable}{3.3}.

@@

				The Development website was not linked correctly; the linked
				hostname did not even exist in DNS, so the URL of the VCS
				could not be determined and the 0.3-r2 version was examined.
				The code is not well modularized (almost everything is to be
				found inside \texttt{include/libratuus.php}), but it is well
				more organized (when it comes to object orientation, systematic
				"routing" and access control for example) than \emph{PostVis
				Admin}. The possibility of \ac{SQL} injections is unclear but
				there is no query sanitization in the
				\texttt{mailClass->as\_db()} function, that makes use of
				\texttt{mysql\_query()}, which in turn supports multiple queries
				configuration dependently. There is an \crit{audit
				trail}{2.4.1} feature but it supports no undo. Also the default
				\crit{password hashing}{3.1} algorithm is \texttt{md5-crypt}.

@@

				\emph{VBox.Adm} makes use of raw \ac{SQL} queries much, but
				sanitizes them. At some places (e.g. in a database maintainance
				script), the queries depend on MySQL. In others (e.g. the
				authentication), they are abstractly noted and \crit{database
				independend}{1.5}. As with most other solutions, there are no
				tests. Solely the custom SMTP proxy for spam maintainance
				integration seems to be tested a little.

@@

				The only PHP tool, which uses an \ac{ORM}. (But migrations were
				not used, yet they are supported by that \ac{ORM}.)
				\emph{ViMbAdmin} has an \crit{audit log}{2.4.1} available but
				it does not support undo. There is no real \crit{SIEVE
				support}{2.4.3}, but an access restriction if configurable for
				mailboxes on different services like SIEVE. The default
				\crit{password hashing}{3.1} algorithm is salted \texttt{MD5}
				(\texttt{md5-crypt}?) and more secure algorithms are only
				supported with shell execution of \texttt{dovecotpw}, which is
				bad for \crit{containability}{3.3}. As with many other compared
				solutions, Twitter bootstrap is used as an \ac{UI} framework,
				which supports responsive design easily but \emph{ViMbAdmin}
				seems not to make use of it conciously (there is no menu on
				mobile phones, for example).

			\begin{landscape}
				\subsubsection*{Comparison overview}
				\label{sec:evaluation:overview}
					\begin{tabular}{l|cccccc|ccccccccc}
						              & 1 & 1.1 & 1.2  & 1.3 & 1.4   & 1.5  & 2 & 2.1 & 2.2 & 2.3  & 2.4 & 2.4.1 & 2.4.2 & 2.4.3 & 2.4.4 \\
						\hline
						modoboa       &   & \y  &  \n  & \y  &  \y   &  \y  &   & \y  & \y  &  \y  &     & (\y)  & \y    &  \y   & \y    \\
						Møil          &   & \y  &  \n  & \y  &  \y   &  \y  &   & \y  & \y  &  \y  &     &  \y   & \n    &  \n   & \n    \\
						postfix.admin &   & \n  &      &     & (\y)  & (\y) &   & \y  & \y  &  \y  &     & (\y)  & \y    &  \n   & \n    \\
						PostVis Admin &   & \n  & (\y) & \n  &  \n   &  \n  &   & \y  & \n  & (\y) &     &  \n   & \n    &  \n   & \y    \\
						ratuus        &   & \n  &  \y  & \y  &  \n   &  \n  &   & \y  & \y  &  \y  &     & (\y)  & \n    &  \n   & \n    \\
						VBox.Adm      &   & \y  &  \n  & \y  &  \n   & (\y) &   & \y  & \y  &  \y  &     &  \n   & \y    &  \n   & \y    \\
						ViMbAdmin     &   & \y  &  \n  & \n  &  \n   &  \y  &   & \y  & \y  &  \y  &     & (\y)  & \n    & (\n)  & \n    \\
						\hline
					\end{tabular}
					\s

					\noindent
					\begin{tabular}{l|cccc|ccc|ccc}
						              & 3 & 3.1  & 3.2  & 3.3  & 4 & 4.1  & 4.2 \\
						\hline
						modoboa       &   &  \y  &  \y  &  \y  &   & (\y) & \y  \\
						Møil          &   &  \y  &  \y  &  \y  &   &  \y  & \y  \\
						postfix.admin &   & (\y) &  \n  & (\y) &   &  \n  & \y  \\
						PostVis Admin &   & (\y) &  \n  &  \n  &   &  \n  & \n  \\
						ratuus        &   & (\y) &  \n  &  \y  &   &  \n  & \n  \\
						VBox.Adm      &   &  \y  & (\y) &  \y  &   &  \n  & \y  \\
						ViMbAdmin     &   & (\y) &  \n  & (\n) &   &  \n  & \n  \\
						\hline
					\end{tabular}
					\s

@@

		That fact alone would maybe suffice as an argument for the development
		for yet another tool but was not as important as some other ideas.
		\emph{Møil} began as a replacement for the database schema shipped with
		\emph{postfix.admin} in a private mail system of mine. The schema was
		intended to be reimplemented normalized \cite{dbnorm} and with a
		possibility of being updated through schema migrations. Also, the
		\ac{Rails} project was intended to provide the basis for command line
		utilities to import the existent data and to perform basic
		administration tasks on domains, mailboxes and aliases. As there has
		been only the schema in use, not the whole \emph{postfix.admin} web
		frontend, there was an extra solution to generate password hashes and
		send them to an administrative contact in place. A replacement for this
		extra tool at first was the only function of the web interface part of
		\emph{Møil}. The migration opportunity was also used to change the
		password hashing algorithm from unsalted \texttt{md5} to
		\texttt{sha512-crypt}, which is substantially more secure
		\cite{missing}. (\texttt{bcrypt} was considered but unfortunately
		rejected, because it is not supported by \texttt{glibc} included in
		Debian Linux 7. Therefore, none of the compared tools supports it
		natively.)

@@

	\newpage
	\section*{Numbered index of criteria}
	\label{sec:appendix:criteria}
		\begin{easylist}
			& Data model
			&& Normalized schema
			&& \emph{postfix.admin} compatible
			&& \emph{postfix.admin} import
			&& Updatable (Migrations)
			&& Systematic database independence
